---
title: "Patient_Selection"
author: "Álvaro Martínez Rubio"
date: "28/10/2024"
output: html_document
---

# Load Libraries 

```{r}
library(flowCore)
library(data.table)
library(ggplot2)
library(readxl)
```

# Output directory

Create a folder to export the merged FCS files of the selected patients

```{r}
destination="Selected"
dir.create(destination)
```

# Input directory

The directory folder should contain one FCS file per patient

```{r}
directory="Merged"
patients=list.files(directory)
```

# Load metadata

Mainly for tracking relapse status

```{r}
database=read_excel("metadata.xlsx")
patients=gsub(".fcs","",patients)
metadata=data.frame(Patient=patients,Status=factor(database$RELAPSE[database$ID%in%patients]))
levels(metadata$Status)=c("NR","R")
```

# Import files
```{r}
patients=paste0(patients,".fcs")
files=list()
for (i in 1:length(patients)){
  files[[i]]=read.FCS(paste0(directory,"/",patients[[i]]))
}
names(files)=patients
```

# Visualize markers per patient (table)

```{r}
markers_list=lapply(files,function(x){unname(x@parameters@data$name)})

# Get the unique list of all possible markers
all_markers = unique(unlist(markers_list))

# Create a matrix to store the presence of markers for each patient
patient_marker_matrix = matrix(0, nrow = length(markers_list), ncol = length(all_markers))

# Assign row and column names
rownames(patient_marker_matrix) = gsub(".fcs","",patients)
colnames(patient_marker_matrix) = all_markers

# Fill the matrix: 1 if marker is present for a patient, otherwise 0
for (i in 1:length(markers_list)) {
  patient_marker_matrix[i, all_markers %in% markers_list[[i]]] = 1
}

# Convert the matrix to a data frame for plotting
patient_marker_df = as.data.frame(patient_marker_matrix)
patient_marker_df$Patient = rownames(patient_marker_df)

# Melt the data for ggplot
melted_df = melt(patient_marker_df, id.vars = "Patient", variable.name = "Marker", value.name = "Presence")
```

```{r}
# Split data according to hospital
subset_hva_hvr = melted_df[grepl("HVA|HVR", melted_df$Patient), ]
subset_other = melted_df[!grepl("HVA|HVR", melted_df$Patient), ]

# Plot HVA and HVR
pdf(paste0(destination, "/mkrpts_HVA_HVR.pdf"), width=10, height=15)
ggplot(subset_hva_hvr, aes(x = Marker, y = Patient, fill = as.factor(Presence))) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("0" = "white", "1" = "black")) +
  theme_bw() +
  labs(x = "Marker", y = "Patient") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position = "none")
dev.off()

# Plot HNJ
pdf(paste0(destination, "/mkrpts_HNJ.pdf"), width=10, height=15)
ggplot(subset_other, aes(x = Marker, y = Patient, fill = as.factor(Presence))) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("0" = "white", "1" = "black")) +
  theme_bw() +
  labs(x = "Marker", y = "Patient") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position = "none")
dev.off()
```


# Remove patients with less than 10000 cells
```{r}
files[which(sapply(files,nrow)<10000)]=NULL

# Recompute markers, number of cells and patient IDs
markers=lapply(files,function(x){unname(x@parameters@data$name)})
ncells=sapply(files,nrow)
patients=names(files)
```

# Extract markers
```{r}
# frequency
freqMarkers=table(unlist(markers))

# list
fullMarkers=names(freqMarkers)
```

# Visualize
```{r}
# Put names and frequencies into dataframe for ggplot
data = data.frame(Marker = names(freqMarkers),Frequency = as.numeric(freqMarkers))

# Retain only markers that are present in more than 100 patients
filtered_data = data[data$Frequency > 100, ]

# Barplot
pdf(paste0(destination,"/freq.pdf"),width=6,height=4.2)
ggplot(filtered_data, aes(x = Marker, y = Frequency)) +
  geom_bar(stat = "identity",fill="darkgray") +
  labs(x = "Marker", y = "Number of patients") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size=10,angle = 90, hjust = 1, vjust=0.5, color = "black"),
    axis.text.y = element_text(size=15, color = "black"),
    axis.title = element_text(size=15,color="black")
  )
dev.off()
```

# Visualize patient-marker tradeoff
```{r}
# List of frequencies to assess
freqs=100:(length(patients)-1)

# Initialize
nmkr=integer()
mkr_selection=list()
npts=integer()
pts_selection=list()

for (i in 1:length(freqs) ){
mkr_selection[[i]]=names(which(freqMarkers>freqs[i])) # Find which markers are present in more than freqs[i] patients
nmkr[i]=length(mkr_selection[[i]]) # count them
pts_selection[[i]]=which(sapply(markers,function(x){all(mkr_selection[[i]] %in% x)})) # find which patients have ALL those markers
npts[i]=length(pts_selection[[i]]) # count them
}

# Put into dataframe for ggplot
df=data.frame(markers=nmkr,patients=npts)

# Plot
pdf(paste0(destination,"/tradeoff.pdf"),width=4,height=4)
ggplot(df,aes(markers,patients))+geom_line(linewidth=2,color="darkgray")+theme_bw()+xlab("Number of markers")+ylab("Number of patients")+theme(axis.text = element_text(size=15,color="black"),axis.title = element_text(size=15,color="black"))
dev.off()
```

# Selection
```{r}
# Specify final selection (number of markers)
number_selection=10

# Subset
marker_selection=mkr_selection[[max(which(nmkr==number_selection))]]
patient_selection=which(sapply(markers,function(x){all(marker_selection %in% x)}))
```

# Export 
```{r}
for (i in patient_selection){
  
  # Export 10000 cells per patient
  ndata=nrow(files[[i]])
  if(ndata>10000){ndata=10000}
  dataf=as.data.frame(files[[i]]@exprs)
  
  if (length(which(marker_selection%in%colnames(dataf)))==length(marker_selection)){
    dataf=dataf[sample(nrow(dataf),ndata),marker_selection] # downsample
    write.FCS(flowFrame(as.matrix(dataf)),paste0(destination,"/",patients[[i]],".fcs")) 
  }
}
```




