---
title: "Patient_Selection"
author: "Álvaro Martínez Rubio"
date: "01/04/2024"
output: html_document
---

# Load Libraries 

```{r}
library(flowCore)
library(data.table)
library(readxl)
```

# Output directory

Create a folder to export the merged FCS files of the selected patients

```{r}
destination="Selected"
dir.create(destination)
```

# Import data

The directory folder should contain one FCS file per patient

```{r}
directory="Merged"
patients=list.files(directory)
```

# Load metadata

Mainly for tracking relapse status

```{r}
database=read_excel("metadata.xlsx")
patients=gsub(".fcs","",patients)
metadata=data.frame(Patient=patients,Status=factor(database$RELAPSE[database$ID%in%patients]))
levels(metadata$Status)=c("NR","R")
```


# Import files
```{r}
files=list()
for (i in 1:length(patients)){
  files[[i]]=read.FCS(paste0(directory,"/",patients[[i]],".fcs"))
}
names(files)=patients
```

# Remove patients with less than 10000 cells
```{r}
files[which(sapply(files,nrow)<10000)]=NULL
markers=lapply(files,function(x){unname(x@parameters@data$name)})
ncells=sapply(files,nrow)
patients=names(files)
```

# Extract markers
```{r}
freqMarkers=table(unlist(markers))
fullMarkers=names(freqMarkers)
```

# Check frequency of markers
```{r}
barplot(freqMarkers,ylab = "Frequency",las=2)
```

# Visualize patient-marker tradeoff
```{r}
freqs=100:(length(patients)-1)
nmkr=integer()
npts=integer()

for (i in 1:length(freqs) ){
nmkr[i]=length(fullMarkers[which(freqMarkers>freqs[i])])
npts[i]=sum(sapply(markers,function(x){all(fullMarkers[which(freqMarkers>freqs[i])] %in% x)}))
}

npts=npts[c(1,which(diff(nmkr)!=0)+1)]
freqs=freqs[c(1,which(diff(nmkr)!=0)+1)]
nmkr=nmkr[c(1,which(diff(nmkr)!=0)+1)]

plot(x=nmkr,y=npts,ylab="Number of patients",xlab="Number of markers",cex=5)
text(x=nmkr,y=npts ,labels=freqs)

```

# Selection
```{r}
marker_selection=c("CD10","CD19","CD20","CD34","CD38","CD45","CD66","CD58", "FSC.A", "SSC.A")
patient_selection=which(sapply(markers,function(x){all(marker_selection %in% x)}))
```

# Export 
```{r}
for (i in patient_selection){
  ndata=nrow(files[[i]])
  if(ndata>10000){ndata=10000}
  dataf=as.data.frame(files[[i]]@exprs)
  if (length(which(marker_selection%in%colnames(dataf)))==length(marker_selection)){
  dataf=dataf[sample(nrow(dataf),ndata),marker_selection] # downsample
  write.FCS(flowFrame(as.matrix(dataf)),paste0(destination,"/",patients[[i]],".fcs")) 
  }
}
```




