---
title: "Visualization"
author: "Álvaro Martínez Rubio"
date: "01/04/2024"
output: html_document
---

# Load Libraries and auxiliary programmes

```{r}
# Flow cytometry
library(flowCore)
library(flownalysis)
library(FlowSOM)

# Data manipulation
library(reshape2)
library(dplyr)
library(ks)
library(uwot)

# Plotting
library(ggplot2)
library(ggcyto)
library(ggpubr)
library(viridis)
library(ggridges)
library(gridExtra)
library(RColorBrewer)

# Excel
library(readxl)
library(writexl)
```

# Output directory

Create a folder to export the visualization results

```{r}
destination="Visualization"
dir.create(destination)
```

# Select patients

The directory folder should contain one FCS file per patient

```{r}
directory="Selected"
patients=list.files(directory,pattern=".fcs")
```

# Load data

```{r}
fn=file.path(directory,patients)
fs=read.flowSet(files=fn)
markers=markernames(fs)
```  

# Load metadata

```{r}
database=read_excel("metadata.xlsx")
patients=gsub(".fcs","",patients)
metadata=data.frame(Patient=patients,Status=factor(database$RELAPSE[database$ID%in%patients]))
levels(metadata$Status)=c("NR","R")
```

# Diagnostic plots

Visualize all markers together with side scatter

```{r}
for (i in 1:length(markers)){
pdf(paste0(destination,"/Diagnostic_",markers[i],".pdf"),width=20,height=22)
print(ggcyto(fs,aes_string(markers[i],"SSC.A"))+geom_hex(bins = 300)+ggcyto_par_set(limits = list(x = c(-0.2,1.2), y= c(-0.2,1.2))))
dev.off()
}
```

# FlowSOM

Concatenate all patients, then cluster

```{r}
all=merge_flowSet(fs,rename.cols = F)
all$exp=gsub(".fcs","",all$exp)
all_ff=flowFrame(as.matrix(all[,markers]))
```

```{r}
fSOM <- FlowSOM(all_ff,xdim = 10, ydim = 5, maxMeta = 20,transform = FALSE,compensate = FALSE,scale = FALSE) 
#load("fsom.Rdata") # alternatively

N=length(levels(fSOM$metaclustering)) # number of metaclusters
cluster=GetClusters(fSOM) # cluster IDs
metacluster=as.numeric(fSOM$metaclustering[cluster]) # metacluster IDs

all_cluster=cbind(all,cluster) # add cluster IDs
relapsecells <- as.data.frame(lapply(metadata, rep, fsApply(fs,nrow))) # all cells concatenated and annotated

save(fSOM,file=paste0(destination,"/fsom.Rdata"))
```

# FlowSOM summary

Plot MST visualization with metacluster ID (p1) and marker expression (p2)

```{r}
colors=c('#8dd3c7','#ffffb3','#bebada','#fb8072','#80b1d3','#fdb462','#b3de69','#fccde5','#d9d9d9','#bc80bd')

pdf(paste0(destination,"/FSOM_Summary.pdf"),width=10,height=5)
p1=PlotVariable(fSOM,variable = fSOM$metaclustering,colorPalette = colors)
ps=PlotStars(fSOM,list_insteadof_ggarrange = TRUE)
p2=ps$tree
ggarrange(p1,p2,nrow=1,ncol=2)
dev.off()
```


# FlowSOM Legend 

Separated for improved visualization

```{r}
pdf(paste0(destination,"/FSOM_legend.pdf"),width=4,height=2)
ps$starLegend
dev.off()
```

# Characterize clusters
```{r}
metacluster_MFI=GetMetaclusterMFIs(fsom=fSOM, colsUsed = FALSE, prettyColnames = FALSE)
metacluster_CV=GetMetaclusterCVs(fsom=fSOM, colsUsed = FALSE, prettyColnames = FALSE)

write_xlsx(metacluster_MFI,paste0(destination,"/MFI.xlsx"))
write_xlsx(metacluster_CV,paste0(destination,"/CV.xlsx"))
```


# Plot marker expression per metacluster

For each metacuster, visualize the aggregated marker expression

```{r}
df=all[]
df$cluster=metacluster
df$condition=metadata$Status[match(df$exp,metadata$Patient)]
meltdf=melt(data.table(df),id=c("exp","condition","cluster"))
meltdf$cluster=factor(meltdf$cluster,levels=c(8,7,6,5,4,3,2,1),ordered = TRUE)
```

```{r}
pdf(paste0(destination,"/clus_exp.pdf"),width=20,height=10)
ggplot(meltdf, aes(x =value, y = cluster,fill=condition)) +
  facet_wrap(~variable, scales = "free_x", nrow = 1) + 
  geom_density_ridges(alpha = 0.2) + scale_fill_manual(values=c("#00BFC4","#F8766D"))+
  theme_ridges()+ 
  theme(legend.position = "none",
        strip.background = element_blank(),
        strip.text = element_text(face = "bold",size=30))+xlim(c(-0.2,1.2))
dev.off()
```

# Plot marker expression per metacluster per patient

Same as above, but without aggregating patient data

```{r}
p=list()
index=1
cols=colnames(all)[-ncol(all)]
for (i in 1:length(levels(metacluster))){
  for (j in cols){
idx=which(metacluster==i)
sub=all[idx,-ncol(all)]
sub$exp=all[idx,ncol(all)]
sub$condition=metadata$Status[match(sub$exp,metadata$Patient)]

color=as.character(metadata[which(metadata$Patient%in%unique(sub$exp)),2])
color[color=="NR"]="#00BFC4"
color[color=="R"]="#F8766D"
p[[index]]=ggplot(sub,aes_string(j,fill="exp"))+geom_density(alpha=0.5)+scale_fill_manual(values=color)+scale_x_continuous(limits=c(-0.2, 1.2),breaks=c(0,0.4,0.8,1.2))+scale_y_continuous(breaks=NULL)+theme_minimal()+theme(legend.position = "none",axis.title = element_blank(),axis.text = element_blank(),axis.ticks = element_blank(),panel.grid.minor.x = element_blank())

index=index+1
}
}
```

```{r}
pdf(paste0(destination,"/clus_exp_pts.pdf"),width=20,height=10)
ggarrange(plotlist=p,nrow=8,ncol=10)
dev.off()
```


# UMAP 

Dimensionality reduction. Package 'uwot'

```{r}
all_mc=cbind(all,metacluster,relapsecells)
all_mc_ds=all_mc[sample(1:nrow(all_mc),1000*length(unique(all_mc$exp))),] # sample 1000 cells per patient
fs_umap=umap(all_mc_ds[,clus_markers],ret_model = TRUE,min_dist = 0.1,n_neighbors = 50)
umap_coord=as.data.frame(fs_umap$embedding)
colnames(umap_coord)=c("UMAP1","UMAP2")
umap_res=cbind(umap_coord,all_mc_ds)
save(umap_res,file=paste0(destination,"/UMAP.RData"))
save_uwot(fs_umap,file=paste0(destination,"/UMAP_model.RData"))
```



# Plot each marker on UMAP embedding
```{r}
plt=list()
for (j in 1:length(markers)){
  plt[[j]]=ggplot(umap_res,aes_string(x="UMAP1",y="UMAP2",color=markers[j]))+geom_point(size=1)+theme_bw()+scale_color_viridis_c(limits = c(-0.2, 1.2),oob = scales::squish,breaks=c(0,0.5,1),labels=c(0,0.5,1))+rremove("grid")+rremove("ticks")+rremove("axis.title")+rremove("axis.text")+ggtitle(markers[j])+theme(panel.border = element_blank(),legend.title = element_blank())+theme(legend.position="none",plot.title = element_text(size=22,face="bold"))
}

pdf(paste0(destination,"/UMAP_mkrs.pdf"),width=22,height=10)
ggarrange(plotlist = plt,nrow=2,ncol=4)
dev.off()
```

# Legend

Separated for improved visualization

```{r}
p=ggplot(umap_res,aes_string(x="UMAP1",y="UMAP2",color=markers[j]))+geom_point(size=1)+theme_bw()+scale_color_viridis_c(limits = c(-0.2, 1.2),oob = scales::squish,breaks=c(0,0.5,1),labels=c(0,0.5,1))+theme(legend.title = element_blank(),legend.position = "bottom")
leg=get_legend(p)
pdf(paste0(destination,"/UMAP_leg.pdf"),width=4,height=1)
as_ggplot(leg)
dev.off()
```


# Plot UMAP with flowSOM metacluster

Color each cell according to metacluster ID

```{r}
pdf(paste0(destination,"/UMAP_mc.pdf"),width=5,height=5)
print(ggplot(umap_res,aes(UMAP1,UMAP2,color=as.factor(metacluster)))+geom_point(size=1)+theme_void()+rremove("grid")+rremove("ticks")+rremove("axis.title")+rremove("axis.text")+theme(legend.position = "none")+scale_color_manual(values=colors))
dev.off()
```






