---
title: "Visualization"
author: "Álvaro Martínez Rubio"
date: "01/04/2024"
output: html_document
---

# Load Libraries and auxiliary programmes

```{r}
# Flow cytometry
library(flowCore)
library(FlowSOM)
library(ConsensusClusterPlus)

# Data manipulation
library(reshape2)
library(dplyr)
library(plyr)
library(ks)
library(uwot)
library(data.table)
library(clue)

# Plotting
library(ggplot2)
library(ggcyto)
library(ggpubr)
library(viridis)
library(ggridges)
library(gridExtra)
library(RColorBrewer)

# Parallel
library(foreach)
library(doParallel)

# Excel
library(readxl)
library(writexl)
```

# Output directory

Create a folder to export the visualization results

```{r}
destination="Visualization"
dir.create(destination)
```

# Select patients

The directory folder should contain one FCS file per patient

```{r}
directory="Selected"
patients=list.files(directory,pattern=".fcs")
```

# Load data

```{r}
fn=file.path(directory,patients)
fs=read.flowSet(files=fn)
markers=markernames(fs)
```  

# Load metadata

```{r}
database=read_excel("metadata.xlsx")
patients=gsub(".fcs","",patients)
metadata=data.frame(Patient=patients,Status=factor(database$RELAPSE[database$ID%in%patients]))
levels(metadata$Status)=c("NR","R")
```

# Diagnostic plots

Visualize all markers versus side scatter

```{r}
for (i in 1:length(markers)){
pdf(paste0(destination,"/Diagnostic_",markers[i],".pdf"),width=20,height=22)
print(ggcyto(fs,aes_string(markers[i],"SSC.A"))+geom_hex(bins = 300)+ggcyto_par_set(limits = list(x = c(-0.2,1.2), y= c(-0.2,1.2))))
dev.off()
}
```

# FlowSOM

Concatenate all patients

```{r}
all=ldply(fsApply(fs, function(x) as.data.frame(exprs(x)), simplify = FALSE),.id = "exp")
all$exp=gsub(".fcs","",all$exp)
all_ff=flowFrame(as.matrix(all[,markers]))
```

Cluster with flowSOM + consensuscluster. Find optimal number of clusters with PAC and area under delta curve. Put all of this in a function and repeat to get estimation of variance.

```{r}
# Function to perform clustering and compute deltaK and PAC
perform_clustering = function(fsom, clus_markers, maxK = 20, reps = 300, pItem = 0.9, pFeature = 1, breaks = 100, x1 = 0.1, x2 = 0.9) {
  
  # Compute Self-Organizing Map
  som = BuildSOM(fsom, colsToUse = which(markers %in% clus_markers), silent = TRUE, xdim = 10, ydim = 5)
  
  # Compute metaclusters
  mc = suppressWarnings(suppressMessages(ConsensusClusterPlus(t(som$map$codes), maxK = maxK, reps = reps, pItem = pItem, pFeature = pFeature,title = destination, plot = "pdf", verbose = FALSE,clusterAlg = "hc", distance = "euclidean", innerLinkage = "complete")))
  
  # vector with numbers of metaclusters
  Kvec = 2:maxK
  
  # initialize
  PAC = rep(NA, length(Kvec))
  names(PAC) = paste("K=", Kvec, sep = "")
  areaK = c()
  
  # for each metaclustering result
  for (i in Kvec) {
    # extract consensus matrix
    M = mc[[i]]$consensusMatrix
    
    # Compute proportion of ambiguously clustered pairs (PAC)
    Fn = ecdf(M[lower.tri(M)])
    PAC[i - 1] = Fn(x2) - Fn(x1)
    
    # Compute area under the CDF (histogram method)
    h = hist(M, plot = FALSE, breaks = seq(0, 1, by = 1 / breaks))
    h$counts = cumsum(h$counts) / sum(h$counts)
    thisArea = 0 # initialize
    for (bi in 1:(length(h$breaks) - 1)) {
      thisArea = thisArea + h$counts[bi] * (h$breaks[bi + 1] - h$breaks[bi])
      bi = bi + 1
    }
    areaK = c(areaK, thisArea)
  }
  
  # Compute change in AUC
  deltaK = areaK[1] # initial auc at k=2
  for (i in 2:(length(areaK))) {
    deltaK = c(deltaK, (areaK[i] - areaK[i - 1]) / areaK[i - 1])
  }
  
  # Put everything into dataframe
  data.frame(Kvec = Kvec, PAC = PAC, deltaAUC = deltaK)
}

# Number of iterations
iterations = 100

# Clustering markers
clus_markers=c("CD45","CD10","CD34","CD19","CD20","CD38","CD58","CD66")

# Read data into FlowSOM
fsom = ReadInput(all_ff,transform = FALSE,compensate = FALSE,scale = FALSE)

# Register parallel backend
cl = makeCluster(detectCores())
registerDoParallel(cl)

# Apply function
results_list = foreach(i = 1:iterations,.packages = c('FlowSOM','ConsensusClusterPlus')) %dopar% {
  perform_clustering(fsom, clus_markers)
}

# Stop the cluster after the task is complete
stopCluster(cl)

# Combine results into one data frame
results = rbindlist(results_list, idcol = "iteration")

# Summarize
summary_results = results[, .(mean_PAC = mean(PAC, na.rm = TRUE), sd_PAC = sd(PAC, na.rm = TRUE), 
                               mean_deltaAUC = mean(deltaAUC, na.rm = TRUE), sd_deltaAUC = sd(deltaAUC, na.rm = TRUE)), by = Kvec]

# Export
write_xlsx(results,path=paste0(destination,"/Clustering.xlsx"))
```

Plot

```{r}
# Plot results with error bars
pdf(paste0(destination,"/Elbow.pdf"),width=4,height=4)
ggplot(summary_results, aes(x = Kvec)) +
  geom_point(aes(y = mean_PAC, color = "PAC"),size=3) +
  geom_line(aes(y = mean_PAC, color = "PAC"),linewidth=1) +
  geom_errorbar(aes(ymin = mean_PAC - sd_PAC, ymax = mean_PAC + sd_PAC, color = "PAC"), width=0.5,linewidth=1) +
  geom_point(aes(y = mean_deltaAUC, color = "deltaAUC"),size=3) +
  geom_line(aes(y = mean_deltaAUC, color = "deltaAUC"),linewidth=1) +
  geom_errorbar(aes(ymin = mean_deltaAUC - sd_deltaAUC, ymax = mean_deltaAUC + sd_deltaAUC, color = "deltaAUC"), width = 0.5,linewidth=1) +
  scale_color_manual(name = "Metric", values = c("PAC" = "#628dbd", "deltaAUC" = "#fa8072"),labels=c("AUC","PAC")) +
  ylab("Value")+xlab("Number of Metaclusters")+theme_bw()+theme(legend.position =c(.75,.75),axis.text = element_text(size=15,color="black"),axis.title = element_text(size=15),legend.text = element_text(size=15),legend.title = element_blank())
dev.off()
```

Tough call. There is a clear k=5 according to AUC. Another at k=8 according to both. A final one at k=15, both again. So if we use both we can argue 8 and 15. (weak)

Metacluster ID may be too specific of one flowSOM run. Let's complicate this by running flowSOM several times and matching metacluster ID by means of consensus clustering in package 'clue'

After deciding optimal number of metaclusters, obtain final metacluster assignment by aggregating results from 50 new runs of FlowSOM 


```{r}
# Number of iterations and clusters
iterations = 50
num_clusters = 9
clus_markers=c("CD45","CD10","CD34","CD19","CD20","CD38","CD58","CD66")

# Store metacluster assignments from each iteration
metacluster_assignments = matrix(NA, nrow = nrow(all_ff@exprs), ncol = iterations)

# Read data into FlowSOM
fsom = ReadInput(all_ff,transform = FALSE,compensate = FALSE,scale = FALSE)

# Number of cores to use
num_cores = detectCores() - 1

# Create a parallel backend
cl = makeCluster(num_cores)
registerDoParallel(cl)

# Run FlowSOM multiple times and store results in parallel
results = foreach(i = 1:iterations, .combine = cbind, .packages = 'FlowSOM') %dopar% {
  
  # SOM
  fsom = BuildSOM(fsom, colsToUse = which(markers %in% clus_markers), silent = TRUE, xdim = 10, ydim = 5)
  
  # Metacluster
  metaClustering = as.character(metaClustering_consensus(fsom$map$codes, k = num_clusters))
  
  # Get metacluster assignments
  GetMetaclusters(fsom, metaClustering)
}

# Stop the cluster
stopCluster(cl)

# Combine results
metacluster_assignments = matrix(results, nrow = nrow(all_ff@exprs), ncol = iterations)

# Use consensus clustering from clue package
metacluster_assignment_clue=lapply(as.list(data.frame(metacluster_assignments)),as.cl_hard_partition)
final_consensus=cl_consensus(metacluster_assignment_clue,method = "HE",control=list(nruns=1))
final_consensus_ids=as.integer(cl_class_ids(final_consensus))

# Save data
mc_IDs=as.data.frame(metacluster_assignments)
save(mc_IDs,file=paste0(destination,"/mc_IDs.RData"))
save(final_consensus_ids,file=paste0(destination,"/consensus_IDs.RData"))
```



# UMAP 

Dimensionality reduction

```{r}
# Bind data with metacluster ID
all_mc=cbind(all,final_consensus_ids) 

# sample for UMAP visualization
sample_ids=sample(1:nrow(all_mc),1000*length(unique(all_mc$exp))) 
all_mc_ds=all_mc[sample_ids,] 

# UMAP (uwot)
fs_umap=umap(all_mc_ds[,which(colnames(all_mc_ds)%in%clus_markers)],ret_model = TRUE,min_dist = 0.01,n_neighbors = 15)

# Store in dataframe
umap_coord=as.data.frame(fs_umap$embedding)
colnames(umap_coord)=c("UMAP1","UMAP2")
umap_res=cbind(umap_coord,all_mc_ds)

# Save results
save(umap_res,file=paste0(destination,"/UMAP.RData"))
umap_file=tempfile("fs_umap")
save_uwot(fs_umap,file=umap_file)
save(fs_umap,file=paste0(destination,"/UMAP_model.RData"))
```

# Plot UMAP with flowSOM metacluster

Color each cell according to metacluster ID

```{r}
colors = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072', '#80b1d3','#fdb462', '#b3de69', '#fccde5', '#d9d9d9','#bc80bd','#ffed6f', '#ff9d00', '#b2e2ff', '#c6d8d0', '#d0a6f0')
pdf(paste0(destination,"/UMAP_mc_robust_consensus.pdf"),width=5,height=5)
print(ggplot(umap_res,aes(UMAP1,UMAP2,color=as.factor(final_consensus_ids)))+geom_point(size=1)+theme_void()+rremove("grid")+rremove("ticks")+rremove("axis.title")+rremove("axis.text")+theme(legend.position = "none")+scale_color_manual(values=colors))
dev.off()
```


# Plot all runs of flowSOM on umap

This is for understanding how consensus select the optimal number of clusters

```{r}
plt=list()

for (j in 1:50){
plt[[j]]=ggplot(umap_res,aes_string("UMAP1","UMAP2",color=as.factor(metacluster_assignments[sample_ids,j])))+geom_point(size=1)+theme_void()+rremove("grid")+rremove("ticks")+rremove("axis.title")+rremove("axis.text")+theme(legend.position = "none")+scale_color_manual(values=colors)
}

png(paste0(destination,"/UMAP_mc_all.png"),width=35,height=50,units="cm",res=300)
ggarrange(plotlist=plt,nrow=9,ncol=6)
dev.off()
```

# Plot each marker on UMAP embedding

```{r}
plt=list()

for (j in 1:length(markers)){
  plt[[j]]=ggplot(umap_res,aes_string(x="UMAP1",y="UMAP2",color=markers[j]))+geom_point(size=1)+theme_bw()+scale_color_viridis_c(limits = c(-0.2, 1.2),oob = scales::squish,breaks=c(0,0.5,1),labels=c(0,0.5,1))+rremove("grid")+rremove("ticks")+rremove("axis.title")+rremove("axis.text")+ggtitle(markers[j])+theme(panel.border = element_blank(),legend.title = element_blank())+theme(legend.position="none",plot.title = element_text(size=22,face="bold"))
}

png(paste0(destination,"/UMAP_mkrs.png"),width=20,height=22,units="cm",res=300)
ggarrange(plotlist = plt,nrow=4,ncol=4)
dev.off()
```

# Legend

Separated for improved visualization

```{r}
p=ggplot(umap_res,aes_string(x="UMAP1",y="UMAP2",color=markers[j]))+geom_point(size=1)+theme_bw()+scale_color_viridis_c(limits = c(-0.2, 1.2),oob = scales::squish,breaks=c(0,0.5,1),labels=c(0,0.5,1))+theme(legend.title = element_blank(),legend.position = "bottom")
leg=get_legend(p)
pdf(paste0(destination,"/UMAP_leg.pdf"),width=4,height=1)
as_ggplot(leg)
dev.off()
```

# Plot number of patients per metacluster

```{r}
# Get unique patients per cluster and merge with metadata to get patient status
patient_cluster_status = all_mc %>%
  distinct(exp, final_consensus_ids) %>%
  inner_join(metadata, by = c("exp" = "Patient"))

pdf(paste0(destination,"/mc_patients.pdf"),width=6.5,height=6.5)
ggplot(patient_cluster_status, aes(x = as.factor(final_consensus_ids), fill = Status)) +
  geom_bar() +
  geom_hline(yintercept = nrow(metadata), linetype = "dashed", color = "black") +
  labs(x = "Metacluster",
       y = "Patients") + 
  scale_fill_manual(values=c("#00BFC4","#F8766D"))+
  theme_bw()+theme(legend.position = "none",text=element_text(size=30),axis.text.x=element_text(size=20))
dev.off()
```



# Visualize marker intensity distributions

```{r}
# add Status information
all_mc$Status=metadata$Status[match(all_mc$exp,metadata$Patient)]

# For each metacluster
for (mc in 1:9){
  
# Filter for each metacluster
all_mc_filtered = all_mc[all_mc$cluster %in% mc, ]
  
# Melt the dataframe for plotting
meltdf = melt(data.table(all_mc_filtered), id.vars = c("exp", "Status", "final_consensus_ids"))
meltdf$cluster = factor(meltdf$cluster, levels = mc, ordered = TRUE)

# Plot aggregated distributions
p1 = ggplot(meltdf, aes(x = value, y = final_consensus_ids, fill = Status)) +
    facet_wrap(~variable, scales = "free_x", nrow = 1) + 
    geom_density_ridges(alpha = 0.6) + 
    scale_fill_manual(values = c("#00BFC4", "#F8766D")) +
    theme_ridges() + 
    theme(
      legend.position = "none",
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", size = 20),
      axis.text.x = element_text(size=15)
    ) + xlim(c(-0.2, 1.2))
  
# Plot segregated by patient
p2 = ggplot(meltdf, aes(x = value, y = final_consensus_ids, fill = Status)) +
    facet_wrap(~variable, scales = "free_x", nrow = 1) + 
    geom_density_ridges(alpha = 0.3, color = "black", aes(group = exp)) +  # Black lines for each patient
    scale_fill_manual(values = c("#00BFC4", "#F8766D")) +
    theme_ridges() + 
    theme(
      strip.background = element_blank(),
      strip.text = element_blank(),
      legend.position = "none",
      axis.text.x = element_text(size=15)
    ) + xlim(c(-0.2, 1.2))
  
pdf(paste0(destination, "/single_clus_exp_",mc_selection,".pdf"), width = 20, height = 5)
print(gridExtra::grid.arrange(p1, p2, nrow = 2))
dev.off()
}
```






