---
title: "Feature Extraction"
output: html_document
date: "2022-12-10"
---

# Load Libraries and auxiliary programmes

```{r}
# Flow cytometry
library(flowCore)
library(flownalysis)
library(FlowSOM)

# Data manipulation
library(reshape)
library(dplyr)
library(data.table)
library(moments)

# Plotting
library(ggplot2)
library(ggpubr)
library(ggridges)
library(scales)
library(viridis)

# Excel
library(readxl)
library(writexl)

source("Utilities/extract_features.R")
```

# Output directory

Create a folder to export the extracted features

```{r}
destination="Features"
dir.create(destination)
```

# Select patients

The directory folder should contain one FCS file per patient

```{r}
directory="Selected"
patients=list.files(directory,pattern=".fcs")
```

# Load data

```{r}
fn=file.path(directory,patients)
fs=read.flowSet(files=fn)
markers=markernames(fs)
```  

# Load metadata

```{r}
database=read_excel("metadata.xlsx")
patients=gsub(".fcs","",patients)
metadata=data.frame(Patient=patients,Status=factor(database$RELAPSE[database$ID%in%patients]))
levels(metadata$Status)=c("NR","R")
```

# Load clustering

```{r}
load("Visualization/fsom.Rdata")
cluster=GetClusters(fSOM)
metacluster=GetMetaclusters(fSOM)

all=merge_flowSet(fs,rename.cols = F)
all$exp=gsub(".fcs","",all$exp)
```


# Extract features

## Abundance
```{r}
dir=paste0(destination,"/Abundance")
dir.create(dir,showWarnings = FALSE)

# For each patient, compute the proportion of their cells that belong to each cluster
abundance=list()
for (i in 1:length(fn)){
 abundance[[i]]=as.integer(table(factor(cluster[which(all$exp==patients[i])],levels=1:length(unique(cluster)))))/nrow(fs[[i]])
}
Abundance=as.data.frame(do.call(rbind, abundance)) 
colnames(Abundance)=1:max(cluster)
Abundance$Status=metadata$Status
rownames(Abundance)=patients

write_xlsx(Abundance,paste0(dir,"/Values.xlsx"))
```

## Plots
```{r}
# Compute the proportion
Abundance_freq=Abundance[,1:max(cluster)]*100

# Compute KStest pvals and print barplot
Kstest_p=apply(Abundance_freq, 2, function(x) ks.test(x[metadata$Status=="R"],x[metadata$Status=="NR"])$p)
  
pdf(paste0(dir,"/KStest_pvals.pdf"),width=50,height=5) 
KS=data.frame(Pval=Kstest_p,Cluster=colnames(Abundance_freq))
KS$Cluster=factor(KS$Cluster, levels = KS$Cluster)
print(ggplot(KS,aes(Cluster,Pval))+geom_col(color="black")+geom_hline(yintercept=0.05, linetype="dashed", color = "red")+theme_bw()+theme(text = element_text(size = 20)))
dev.off()

# Boxplot (R vs NR)
Abundance_freq$Status=Abundance$Status

pdf(paste0(dir,"/Boxplot.pdf"),width=10,height=5)
p1=ggplot(melt(data.table(Abundance_freq[,c(1:25,51)]),id="Status"), aes(x=factor(variable), y=value,color=Status))+geom_boxplot(outlier.colour = NULL)+ylab("")+xlab("")+theme_bw()+theme(legend.position = "top")+scale_color_manual(values=c("#00BFC4","#F8766D"))+
  scale_y_continuous(trans=asinh_trans())
p2=ggplot(melt(data.table(Abundance_freq[,26:51]),id="Status"), aes(x=factor(variable), y=value,color=Status))+geom_boxplot(outlier.colour = NULL)+ylab("")+xlab("")+theme_bw()+theme(legend.position = "top")+scale_color_manual(values=c("#00BFC4","#F8766D"))+
  scale_y_continuous(trans=asinh_trans())
ggarrange(p1,p2,ncol=1,nrow=2)
dev.off() 

# Plot pvalue of ks test on flowsom
pdf(paste0(dir,"/fSOM_pval.pdf"),width=7,height=5)
PlotVariable(fSOM,variable=Kstest_p,colorPalette = viridis,lim = c(0,0.05),backgroundValues=fSOM$metaclustering,backgroundColor=colors)
dev.off()

# Plot cluster number
pdf(paste0(dir,"/fSOM_num.pdf"),width=7,height=5)
PlotNumbers(fSOM,backgroundValues=fSOM$metaclustering,backgroundColor=colors)
dev.off()
```


## Abundance (only those with significant differences)
```{r}
sig=which(Kstest_p<0.05)

if (length(sig)>0){
    
dir=paste0(destination,"/Abundance_sig")
dir.create(dir,showWarnings = FALSE)

Abundance_sig=Abundance[sig]

Abundance_sig_freq=Abundance_sig*100
  
# Compute KStest pvals and print barplot
Kstest_p=apply(Abundance_sig_freq, 2, function(x) ks.test(x[metadata$Status=="R"],x[metadata$Status=="NR"])$p)
  
pdf(paste0(dir,"/KStest_pvals.pdf"),width=50,height=5) 
KS=data.frame(Pval=Kstest_p,Marker=colnames(Abundance_sig_freq))
KS$Marker=factor(KS$Marker, levels = KS$Marker)
print(ggplot(KS,aes(Marker,Pval))+geom_col(color="black")+geom_hline(yintercept=0.05, linetype="dashed", color = "red")+theme_bw()+theme(text = element_text(size = 20)))
dev.off()

Abundance_sig_freq$Status=Abundance$Status

pdf(paste0(dir,"/Boxplot.pdf"),width=3,height=5)
print(ggplot(melt(data.table(Abundance_sig_freq),id="Status"), aes(x=factor(variable), y=value,color=Status))+geom_boxplot(outlier.colour = NULL)+ylab("")+xlab("")+theme_bw()+theme(legend.position = "top")+scale_color_manual(values=c("#00BFC4","#F8766D"))+scale_y_continuous(trans=asinh_trans()))
dev.off() 

Abundance_sig$Status=Abundance$Status
      
write_xlsx(Abundance_sig,paste0(dir,"/Values.xlsx"))

}

```


## Expression  (whole dataset)
```{r}
dir=paste0(destination,"/Full")
dir.create(dir,showWarnings = FALSE)

extract_features(fs,metadata,dir)
```

## Expression  (each metacluster)
```{r}
for (i in 1:length(levels(metacluster))){
  
  # Create folder
  dir=paste0(destination,"/Meta_",i)
  dir.create(dir,showWarnings = FALSE)
  
  # Get flowSet
  idx=which(metacluster==i)
  sub=all[idx,-ncol(all)]
  sub_split=split(sub,all[idx,ncol(all)])
  sub_split_ff=lapply(sub_split,function (x) flowFrame(as.matrix(x)))
  sub_fs=flowSet(sub_split_ff)
  
  # Get metadata
  sub_metadata=metadata[which(names(sub_fs@frames)%in%metadata$Patient),]
  
  # Extract features
  extract_features(sub_fs,sub_metadata,dir)
  
}
```










