---
title: "Merge Quality"
author: "Álvaro Martínez Rubio"
date: "2023-07-05"
output: html_document
---

```{r}
library(ggplot2)
library(readxl)
library(data.table)
library(ggsci)
library(ggpubr)
library(scales)
library(flowCore)
library(viridis)
library(writexl)
library(foreach)
library(doParallel)

source("EMD_fcs.R")
```

# Select patients before merging

```{r}
directory_pre="Preprocessed"
patients_pre=list.files(directory_pre)
```

# Select merged patients
```{r}
directory_basic="Merged_basic"
directory_cycombine="Merged_cycombine"
directory_cytobb="Merged_cytobb"
directory_cytofmerge="Merged_cytofmerge"
```

# Read files for each patient and compute EMD

```{r}
EMDs=foreach(i=1:length(patients_pre))%dopar%{
  EMD_basic=EMD_fcs(paste0(directory_pre,"/",patients_pre[i]),paste0(directory_basic,"/",patients_pre[i],".fcs"))
  EMD_cycombine=EMD_fcs(paste0(directory_pre,"/",patients_pre[i]),paste0(directory_cycombine,"/",patients_pre[i],".fcs"))
  EMD_cytobb=EMD_fcs(paste0(directory_pre,"/",patients_pre[i]),paste0(directory_cytobb,"/",patients_pre[i],".fcs"))
  EMD_cytofmerge=EMD_fcs(paste0(directory_pre,"/",patients_pre[i]),paste0(directory_cytofmerge,"/",patients_pre[i],".fcs"))
  EMDs=rbindlist(list(basic=EMD_basic,cycombine=EMD_cycombine,cytobb=EMD_cytobb,cytofmerge=EMD_cytofmerge),idcol="method")
}
names(EMDs)=patients_pre
```

# Prepare data for visualization

```{r}
# Bind
EMD=rbindlist(EMDs,fill=TRUE,idcol="Patient")

# Select only markers that are present in more than 50% of patients
EMD_filtered = EMD[, colSums(is.na(EMD)) / nrow(EMD) < 0.5]

# Remove backbone markers
EMD_filtered = EMD_filtered[, !(colnames(EMD_filtered) %in% c("CD19", "CD45", "CD34"))]

# Melt
meltData=melt(EMD_filtered,id=c("method","Patient","Cells_pre","Cells_post","Time"))
```

# Boxplot of each marker, each method, log scale

```{r}
colors=c("#E48440","#459AB0","#DA3831","#634B8E")

meltData$method=factor(meltData$method,levels=c("basic","cytofmerge","cytobb","cycombine"))

meltData_edit = na.omit(replace(meltData, meltData == 0, 1e-6))

p1=ggplot(meltData_edit[1:(0.5*nrow(meltData_edit)),], aes(x=variable, y=value,fill=method))+geom_boxplot()+ylab(paste0("EMD"))+theme_bw()+xlab("")+ guides(fill=guide_legend(title="Method"))+theme(legend.position = "none",axis.text = element_text(color="black",size=11))+
  scale_y_log10(limits=c(1e-3,5),labels = trans_format("log10", math_format(10^.x)))+scale_fill_manual(values = colors)

p2=ggplot(meltData_edit[(0.5*nrow(meltData_edit)+1):nrow(meltData_edit),], aes(x=variable, y=value,fill=method))+geom_boxplot()+ylab(paste0("EMD"))+theme_bw()+xlab("")+guides(fill=guide_legend(title="Method"))+theme(legend.position = "none",axis.text = element_text(color="black",size=11))+
  scale_y_log10(limits=c(1e-3,5),labels = trans_format("log10", math_format(10^.x)))+scale_fill_manual(values = colors)

# Arrange and export
pdf("/Users/alvaromartinezrubio/Desktop/Merging_quality_log.pdf",width=8,height=4)
ggarrange(p1, p2, nrow = 2)
dev.off()
```


# Visual comprobation

Show the difference between the original and the four imputations, for all imputed markers

```{r}

for (pt in patients_pre){

  # Read files without merging
  filenames=list.files(paste0(directory_pre,"/",pt),pattern = ".fcs",full.names = TRUE)
  files=list()
  for (i in 1:length(filenames)){
    # Read FCS
    data=read.FCS(filenames[i],truncate_max_range = FALSE,column.pattern = "Time",invert.pattern = TRUE) # Read
    files[[i]]=as.data.frame(data@exprs)
  }
  markers=lapply(files,colnames)
  Original=rbindlist(files,fill=TRUE)
  
  # Read merged files
  Merged_basic=read.FCS(paste0(directory_basic,"/",pt,".fcs"))
  Merged_basic=as.data.frame(Merged_basic@exprs)
  
  Merged_cycombine=read.FCS(paste0(directory_cycombine,"/",pt,".fcs"))
  Merged_cycombine=as.data.frame(Merged_cycombine@exprs)
  
  Merged_cytobb=read.FCS(paste0(directory_cytobb,"/",pt,".fcs"))
  Merged_cytobb=as.data.frame(Merged_cytobb@exprs)
  
  Merged_cytofmerge=read.FCS(paste0(directory_cytofmerge,"/",pt,".fcs"))
  Merged_cytofmerge=as.data.frame(Merged_cytofmerge@exprs)

  # Initialize
  plot_list = list()

  # Loop over each marker 
  
  selection = setdiff(colnames(Merged_basic), c("CD45", "CD19", "CD34","FSC.A","SSC.A","FSC.H","SSC.H"))
  
  for (marker in selection) {
    
  # Select sample size (minimum number of cell across method)
  samp=min(length(na.omit(Original[[marker]])),nrow(Merged_basic),nrow(Merged_cycombine),nrow(Merged_cytobb),nrow(Merged_cytofmerge))
  
  # Sample
  marker_original = na.omit(Original[[marker]])
  marker_original = marker_original[sample(1:length(marker_original),samp)]
  marker_basic = Merged_basic[sample(1:nrow(Merged_basic),samp),marker]
  marker_cycombine = Merged_cycombine[sample(1:nrow(Merged_cycombine),samp), marker]
  marker_cytobb = Merged_cytobb[sample(1:nrow(Merged_cytobb),samp), marker]
  marker_cytofmerge = Merged_cytofmerge[sample(1:nrow(Merged_cytofmerge),samp),marker]
    
  # Combine 
  df_combined = data.frame(
    Value = c(marker_original,  marker_basic,marker_cytofmerge, marker_cytobb,  marker_cycombine),
    Source = factor(c(rep("files", length(marker_original)),
                      rep("Merged_basic", length(marker_basic)),
                      rep("Merged_cytofmerge", length(marker_cytofmerge)),
                      rep("Merged_cytobb", length(marker_cytobb)),
                      rep("Merged_cycombine", length(marker_cycombine))),levels=c("files","Merged_basic","Merged_cytofmerge","Merged_cytobb","Merged_cycombine"))
  )
    
  # Density plot 
  p = ggplot(df_combined, aes(x = Value, color = Source)) +
    geom_density(data = subset(df_combined, Source == "files"), aes(color = Source, fill = Source), alpha = 0.5, linewidth=1.5) +
    geom_density(data = subset(df_combined, Source != "files"), aes(color = Source), alpha=0.2,fill = NA, linewidth=0.7) +
    scale_color_manual(values = c("black", colors)) +   
    scale_fill_manual(values = c("black", colors)) +
    theme_bw() + scale_x_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+xlab(marker) +ylab("")+theme(legend.position = "none")
    
    # Add the plot to the list
    plot_list[[marker]] = p
  
}

  # Arrange and plot
  pdf(paste0("Merging_densities_",pt,".pdf"),width=3,height=10) 
  print(ggarrange(plotlist = plot_list, nrow = 9, ncol = 2))
  dev.off()

}

```

Scatter plots comparing backbone with a few relevant markers, for an example patient

```{r}

pt=patients_pre[1]

# Read files without merging
filenames=list.files(paste0(directory_pre,"/",pt),pattern = ".fcs",full.names = TRUE)
files=list()
for (i in 1:length(filenames)){
  # Read FCS
  data=read.FCS(filenames[i],truncate_max_range = FALSE,column.pattern = "Time",invert.pattern = TRUE) # Read
  files[[i]]=as.data.frame(data@exprs)
}
markers=lapply(files,colnames)
Original=rbindlist(files,fill=TRUE)
  
# Read merged files
Merged_basic=read.FCS(paste0(directory_basic,"/",pt,".fcs"))
Merged_basic=as.data.frame(Merged_basic@exprs)

Merged_cycombine=read.FCS(paste0(directory_cycombine,"/",pt,".fcs"))
Merged_cycombine=as.data.frame(Merged_cycombine@exprs)
 
Merged_cytobb=read.FCS(paste0(directory_cytobb,"/",pt,".fcs"))
Merged_cytobb=as.data.frame(Merged_cytobb@exprs)
 
Merged_cytofmerge=read.FCS(paste0(directory_cytofmerge,"/",pt,".fcs"))
Merged_cytofmerge=as.data.frame(Merged_cytofmerge@exprs)
  
# Initialize
plot_list = list()

# Select markers
bbmarker=c("CD45","CD19","CD34")
selection=c("CD10","CD20","CD38")

# Loop over each marker
for (i in 1:length(bbmarker)) {
  
    bb=bbmarker[i]
    sel=selection[i]
    n=length(na.omit(Original[[sel]]))
    
    p1 = ggplot(Original, aes_string(x = bb, y=sel)) + geom_hex(bins=50)+
      theme_bw() + scale_x_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+scale_y_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+xlab("")+ylab("")+
      theme(legend.position = "none",axis.text = element_text(color="black",size=12)) +
  scale_fill_viridis_c()
    
    p2 = ggplot(Merged_basic[sample(1:nrow(Merged_basic),n),], aes_string(x = bb, y=sel)) + geom_hex(bins=50)+
      theme_bw() + scale_x_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+scale_y_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+xlab("")+ylab("")+
      theme(legend.position = "none",axis.text = element_text(color="black",size=12)) +
  scale_fill_viridis_c()
        
    p3 = ggplot(Merged_cytofmerge[sample(1:nrow(Merged_cycombine),n),], aes_string(x = bb, y=sel)) + geom_hex(bins=50)+
      theme_bw() + scale_x_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+scale_y_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+xlab("")+ylab("")+
      theme(legend.position = "none",axis.text = element_text(color="black",size=12)) +
  scale_fill_viridis_c()
    
    p4 = ggplot(Merged_cytobb[sample(1:nrow(Merged_basic),n),], aes_string(x = bb, y=sel)) + geom_hex(bins=50)+
      theme_bw() + scale_x_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+scale_y_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+xlab("")+ylab("")+
      theme(legend.position = "none",axis.text = element_text(color="black",size=12)) +
  scale_fill_viridis_c()
        
    p5 = ggplot(Merged_cycombine[sample(1:nrow(Merged_cycombine),n),], aes_string(x = bb, y=sel)) + geom_hex(bins=50)+
      theme_bw() + scale_x_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+scale_y_continuous(breaks=c(0,1),limits=c(-0.2,1.2))+xlab("")+ylab("")+
      theme(legend.position = "none",axis.text = element_text(color="black",size=12)) +
  scale_fill_viridis_c()
    
    # Add to the list
    plot_list[[i]] = ggarrange(p1, p2, p3, p4, p5, nrow = 1, ncol = 5,align = "hv", widths = c(1, 1, 1, 1, 1)) 
  
}

# Arrange and export
pdf(paste0("Merging_2d_",pt,".pdf"),width=8,height=5)
ggarrange(plotlist = plot_list, nrow = 3, ncol = 1, align = "hv",heights = c(-1, -1, -1)) 
dev.off()

```

# Select a single method and plot 2d for all patients

```{r}
fcs_files = list.files(directory_cytobb, full.names = TRUE)

# Initialize 
plot_list_cd10_cd45 = list()
plot_list_cd20_cd19 = list()
plot_list_cd38_cd34 = list()

# Loop through each FCS file and create the scatter plots
for (i in 1:length(fcs_files)) {
  
  # Read FCS file
  fcs_data = read.FCS(fcs_files[i])
  
  # Extract patient IDs
  patient_id = tools::file_path_sans_ext(basename(fcs_files[i]))
  
  # Convert to dataframe
  fcs_df = as.data.frame(exprs(fcs_data))
  fcs_df = fcs_df[sample(1:nrow(fcs_df),min(10000,nrow(fcs_df))),]
  
    if (all(c("CD10", "CD45", "CD20", "CD19", "CD38", "CD34") %in% colnames(fcs_df))) {

    p1 = ggplot(fcs_df, aes(x = CD45, y = CD10)) +
      geom_hex(bins=80)  + xlab("")+ ylab("")+scale_x_continuous(limits=c(-0.2,1.2),breaks=c(0,1))+scale_y_continuous(limits=c(-0.1,1.1),breaks=c(0,1))+
      ggtitle(patient_id) + scale_fill_viridis_c()+ 
      theme_minimal()+ theme(legend.position = "none")
    
    p2 = ggplot(fcs_df, aes(x = CD19, y = CD20)) +
      geom_hex(bins=80) + xlab("")+ ylab("")+scale_x_continuous(limits=c(-0.2,1.2),breaks=c(0,1))+scale_y_continuous(limits=c(-0.1,1.1),breaks=c(0,1))+
      ggtitle(patient_id)+ scale_fill_viridis_c()+
      theme_minimal()+ theme(legend.position = "none")
    
    p3 = ggplot(fcs_df, aes(x = CD34, y = CD38)) +
       geom_hex(bins=80) + xlab("")+ ylab("")+scale_x_continuous(limits=c(-0.2,1.2),breaks=c(0,1))+scale_y_continuous(limits=c(-0.1,1.1),breaks=c(0,1))+
      ggtitle(patient_id) + scale_fill_viridis_c()+ 
      theme_minimal()+ theme(legend.position = "none")
    
    # Store
    plot_list_cd10_cd45[[i]] = p1
    plot_list_cd20_cd19[[i]] = p2
    plot_list_cd38_cd34[[i]] = p3
    }
}

# Plot all patients on single pages for each comparison

# CD10 vs CD45
pdf("CD10_CD45.pdf",width=22,height=20)
ggarrange(plotlist = plot_list_cd10_cd45, ncol = 11, nrow = 10)
dev.off()

# CD20 vs CD19
pdf("CD20_CD19.pdf",width=22,height=20)
ggarrange(plotlist = plot_list_cd20_cd19, ncol = 11, nrow = 10)
dev.off()

# CD38 vs CD34
pdf("CD38_CD34.pdf",width=22,height=20)
ggarrange(plotlist = plot_list_cd38_cd34, ncol = 11, nrow = 10)
dev.off()

```





